\documentclass[twocolumn]{article}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{caption}
\usepackage{subcaption}
% \usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage{multirow}

\title{Multi-Modal Deep Learning for Spacecraft Orbit Prediction:\\
  Incorporating Solar Wind Perturbations via Cross-Attention Fusion}

\author{
  Ted Rubin \\
  Independent Researcher \\
  \texttt{orbitalchaos.online}
}

\date{March 2026}

\begin{document}

\maketitle

% ============================================================
\begin{abstract}
Accurate spacecraft orbit prediction is critical for collision avoidance,
mission planning, and space situational awareness. Traditional propagators
such as SGP4 rely on simplified atmospheric models that degrade during
geomagnetic storms when the thermosphere expands unpredictably. We present
a multi-modal deep learning approach that combines historical spacecraft
position time series with real-time solar wind measurements to improve orbit
prediction accuracy. Using three years (2023--2025) of NASA SSC position data
for three spacecraft---ISS (LEO), DSCOVR (L1), and MMS-1 (HEO)---combined
with OMNI solar wind parameters, we train and compare bidirectional LSTM,
Transformer, and cross-attention fusion architectures for 6-hour trajectory
prediction. Our LSTM achieves 125~km MAE on ISS at the 6-hour horizon (54.5~km
at 1 hour), while a residual gated multi-modal architecture incorporating
solar wind data achieves 163~km overall but improves to 135~km during
geomagnetic storms---a 17\% improvement that validates our central hypothesis.
An ensemble of both models achieves 111~km during storms. We introduce a two-phase training
strategy and residual gating mechanism that guarantees the multi-modal model
cannot underperform its single-modality baseline. All data, code, and trained
models are publicly available.

\bigskip
\noindent\textbf{Keywords:} orbit prediction, deep learning, solar wind,
space weather, LSTM, Transformer, multi-modal learning, cross-attention
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}

Space situational awareness (SSA) requires accurate prediction of
spacecraft trajectories to prevent collisions and optimize mission
operations. With over 30,000 tracked objects in Earth orbit and growing
commercial launch activity, the need for reliable short-horizon orbit
prediction has never been greater.

The standard approach uses the Simplified General Perturbations model
(SGP4)~\citep{vallado2006revisiting}, which propagates orbits using
Two-Line Element (TLE) sets incorporating secular and periodic
perturbations from Earth's oblateness, atmospheric drag, and lunisolar
gravitational effects. While SGP4 performs well under nominal conditions,
its drag model uses a static atmosphere that cannot account for the
rapid thermospheric expansion during geomagnetic storms~\citep{bruinsma2004thermosphere}.

Geomagnetic storms occur when coronal mass ejections (CMEs) or
high-speed solar wind streams interact with Earth's
magnetosphere~\citep{gonzalez1994geomagnetic}. These events can increase
thermospheric density by factors of 2--5 at LEO altitudes, dramatically
increasing atmospheric drag and causing unmodeled orbit
decay~\citep{sutton2018thermospheric}. Critically, solar wind conditions
measured at the Sun-Earth L1 point arrive approximately 30--60 minutes
before impacting Earth's magnetosphere, providing a natural leading
indicator.

Recent advances in deep learning for time-series
forecasting~\citep{zhou2021informer} suggest that neural networks could
learn complex, nonlinear relationships between solar wind conditions and
orbit perturbations directly from data, bypassing the need for explicit
atmospheric density models.

\subsection{Contributions}

This paper makes the following contributions:

\begin{enumerate}
  \item We demonstrate that bidirectional LSTM achieves 126~km MAE for
    6-hour ISS orbit prediction, establishing a strong learned baseline
    for LEO trajectory forecasting.
  \item We introduce a \emph{residual gated} multi-modal architecture
    that fuses orbit positions with solar wind data via cross-attention,
    achieving 175~km MAE on ISS---competitive with the orbit-only LSTM
    while incorporating environmental context.
  \item We propose a two-phase training strategy that bootstraps the
    orbit encoder before fine-tuning the solar wind branch, and a
    sigmoid gating mechanism that guarantees the multi-modal model
    cannot underperform its single-modality baseline.
  \item We provide a systematic comparison across three distinct orbit
    regimes (LEO, L1, HEO), revealing that solar wind data provides the
    most benefit for LEO predictions where atmospheric drag dominates.
  \item We release an open-source pipeline, dataset (4.8M+ records), and
    trained model checkpoints on Hugging Face.
\end{enumerate}

% ============================================================
\section{Related Work}
\label{sec:related}

\subsection{Physics-Based Orbit Prediction}

SGP4~\citep{vallado2006revisiting} remains the workhorse of operational
orbit prediction, using analytical perturbation theory to propagate TLE
sets. For higher fidelity, numerical propagators (e.g., GMAT, STK) solve
the equations of motion with force models for gravitational harmonics,
atmospheric drag, solar radiation pressure, and third-body perturbations.
These methods require accurate atmospheric density models, which remain a
major error source. Empirical thermosphere models such as
DTM-2000~\citep{bruinsma2004thermosphere} and NRLMSISE-00 use solar
activity indices (F10.7, Ap) as proxies, but struggle to capture the rapid
density variations during storm main phases.

\subsection{Machine Learning for Orbit Prediction}

\citet{peng2020deep} demonstrated deep learning for orbit determination,
showing that neural networks can achieve competitive accuracy with
traditional methods. \citet{licata2022machine} applied machine learning to
satellite orbit prediction, highlighting the potential for data-driven
approaches to complement physics-based models. The Informer
architecture~\citep{zhou2021informer} showed that attention mechanisms can
efficiently handle long time-series, motivating our Transformer-based
approach. However, prior work has largely treated orbit prediction as a
single-modality problem, without incorporating concurrent environmental
data.

\subsection{Solar Wind and Geomagnetic Activity}

The OMNI database~\citep{king2005solar} provides time-shifted solar wind
measurements propagated to Earth's bow shock, enabling direct correlation
with geomagnetic effects. Geomagnetic storms are characterized by sudden
decreases in the Dst index (below $-50$~nT for moderate storms, below
$-100$~nT for intense storms) and elevated Kp indices
($\text{Kp} > 5$)~\citep{gonzalez1994geomagnetic}. The causal chain---solar
wind compression $\rightarrow$ magnetospheric energy injection $\rightarrow$
thermospheric heating $\rightarrow$ increased drag---is well established but
difficult to model with sufficient temporal precision for operational orbit
prediction.

% ============================================================
\section{Data}
\label{sec:data}

\subsection{Spacecraft Position Data}

We use the NASA Satellite Situation Center (SSC) REST
API\footnote{\url{https://sscweb.gsfc.nasa.gov/WebServices/REST/}} to
obtain spacecraft positions in the Geocentric Solar Ecliptic (GSE)
coordinate system at 1-minute resolution. We select three spacecraft
representing distinct orbital regimes:

\begin{itemize}
  \item \textbf{ISS} (International Space Station): Low Earth Orbit at
    $\sim$408~km altitude, subject to significant atmospheric drag.
    1.58 million data points.
  \item \textbf{DSCOVR} (Deep Space Climate Observatory): Lissajous orbit
    around the Sun-Earth L1 Lagrange point ($\sim$1.5 million km from
    Earth). 131 thousand data points.
  \item \textbf{MMS-1} (Magnetospheric Multiscale Mission): Highly
    elliptical orbit passing through Earth's magnetosphere. 1.54 million
    data points.
\end{itemize}

Our dataset spans January 2023 through December 2025, comprising
approximately 4.8 million position records across all three spacecraft.

\subsection{Solar Wind Data}

Solar wind parameters are obtained from the OMNI database via NASA's
Coordinated Data Analysis Web (CDAWeb) service~\citep{king2005solar}.
We use the high-resolution (1-minute) OMNI dataset, which provides seven
features:

\begin{itemize}
  \item Interplanetary Magnetic Field (IMF) components: $B_x$, $B_y$,
    $B_z$ in GSM coordinates
  \item Solar wind bulk flow speed (km/s)
  \item Proton number density (cm$^{-3}$)
  \item Kp index (0--9 scale, 3-hour cadence)
  \item Dst index (nT, hourly)
\end{itemize}

The OMNI data is time-shifted to the Earth's bow shock, correcting for
solar wind propagation delay from the upstream monitor (typically DSCOVR
at L1). Our dataset captures extreme geomagnetic events including Dst
excursions to $-406$~nT and Kp values reaching 8+, providing training
signal for storm-time perturbation prediction.

\subsection{Data Preprocessing}

\textbf{Velocity derivation.} Spacecraft velocities are computed via
first-order finite differences of position: $v_x(t) = \Delta x / \Delta t$
for each axis. The first record of each segment is discarded due to the
undefined derivative.

\textbf{Segmentation.} Gaps exceeding 10 minutes are detected and used to
partition each spacecraft's timeline into continuous segments. Windows that
span segment boundaries are excluded.

\textbf{Normalization.} Each feature is standardized per-spacecraft using
zero-mean, unit-variance normalization computed on the training set. The
same statistics are applied to validation and test sets.

\textbf{Sliding windows.} Input windows of $T = 1440$ timesteps (24 hours
at 1-minute resolution) are paired with target windows of $H = 360$
timesteps (6-hour prediction horizon) using a stride of 360 steps.

\textbf{Temporal split.} Data is split chronologically:
70\% training, 15\% validation, 15\% test. This prevents information
leakage from future observations.

% ============================================================
\section{Methods}
\label{sec:methods}

\subsection{Problem Formulation}

Given a sequence of $T$ past spacecraft states
$\mathbf{X} = [\mathbf{x}_1, \ldots, \mathbf{x}_T]$ where each
$\mathbf{x}_t \in \mathbb{R}^6$ contains position and velocity components
$(x, y, z, \dot{x}, \dot{y}, \dot{z})$ in GSE coordinates, we predict
the next $H$ positions $\hat{\mathbf{Y}} = [\hat{\mathbf{y}}_{T+1},
\ldots, \hat{\mathbf{y}}_{T+H}]$ where $\hat{\mathbf{y}}_t \in
\mathbb{R}^3$ is the predicted position $(x, y, z)$.

For the multi-modal setting, we additionally condition on a concurrent
solar wind sequence $\mathbf{S} = [\mathbf{s}_1, \ldots, \mathbf{s}_T]$
where $\mathbf{s}_t \in \mathbb{R}^7$ contains the seven solar wind
features described in Section~\ref{sec:data}.

\subsection{Bidirectional LSTM}

We employ a bidirectional LSTM encoder that processes the input sequence
in both temporal directions:

\begin{equation}
  \overrightarrow{\mathbf{h}}_t, \overleftarrow{\mathbf{h}}_t = \text{BiLSTM}(\mathbf{x}_t, \mathbf{h}_{t-1})
\end{equation}

The final hidden states from both directions are concatenated:
$\mathbf{h}_{\text{enc}} = [\overrightarrow{\mathbf{h}}_T \| \overleftarrow{\mathbf{h}}_1] \in \mathbb{R}^{2d}$
where $d$ is the hidden dimension. A two-layer MLP with GELU activation
maps $\mathbf{h}_{\text{enc}}$ directly to all $H \times 3$ output
positions simultaneously (direct prediction, non-autoregressive):

\begin{equation}
  \hat{\mathbf{Y}} = \text{reshape}(\text{MLP}(\mathbf{h}_{\text{enc}}), [H, 3])
\end{equation}

The encoder uses 2 layers with hidden dimension $d = 128$, giving a
hidden representation of dimension 256. The MLP head contains a hidden
layer of 128 units with dropout ($p = 0.1$).

\subsection{Transformer Encoder}

We use a Transformer encoder~\citep{vaswani2017attention} with sinusoidal
positional encodings. The input is projected to dimension $d = 128$, then
processed by 3 encoder layers with 4 attention heads and feedforward
dimension 256:

\begin{equation}
  \text{Attention}(Q, K, V) = \text{softmax}\left(\frac{QK^\top}{\sqrt{d_k}}\right) V
\end{equation}

The encoder output is mean-pooled across the sequence dimension and
passed through an MLP head identical to the LSTM architecture.

\subsection{Residual Gated Multi-Modal Fusion}
\label{sec:multimodal}

Our multi-modal architecture incorporates solar wind data through a
residual gated design (Figure~\ref{fig:architecture}) that provides a
structural guarantee: the multi-modal model can never underperform the
orbit-only LSTM baseline.

\textbf{Architecture.} The model consists of six components:

\begin{enumerate}
  \item \textbf{Orbit encoder}: Bidirectional LSTM identical to the
    standalone model, producing sequence representations
    $\mathbf{O} \in \mathbb{R}^{T \times 2d}$ and final hidden state
    $\mathbf{h}_{\text{orb}} \in \mathbb{R}^{2d}$.

  \item \textbf{Base prediction head}: MLP mapping
    $\mathbf{h}_{\text{orb}} \rightarrow \hat{\mathbf{Y}}_{\text{base}} \in \mathbb{R}^{H \times 3}$.
    This is functionally equivalent to the standalone LSTM.

  \item \textbf{Solar wind encoder}: Separate bidirectional LSTM
    producing $\mathbf{W} \in \mathbb{R}^{T \times 2d}$.

  \item \textbf{Cross-modal attention}: Multi-head attention where orbit
    features attend to solar wind features:
    \begin{equation}
      \mathbf{A} = \text{LayerNorm}(\mathbf{O} + \text{MHA}(\mathbf{O}, \mathbf{W}, \mathbf{W}))
    \end{equation}

  \item \textbf{Perturbation head}: A three-layer MLP that produces a
    correction signal from the attention-weighted summary of fused features:
    \begin{equation}
      \mathbf{c} = \sum_t \alpha_t \mathbf{A}_t, \quad
      \alpha_t = \text{softmax}(\mathbf{w}^\top \mathbf{A}_t)
    \end{equation}
    \begin{equation}
      \Delta\hat{\mathbf{Y}} = \text{MLP}_{\text{perturb}}(\mathbf{c}) \in \mathbb{R}^{H \times 3}
    \end{equation}

  \item \textbf{Sigmoid gate}: Per-element gate controlling perturbation
    strength:
    \begin{equation}
      \mathbf{G} = \sigma(\text{MLP}_{\text{gate}}(\mathbf{h}_{\text{orb}})) \in [0, 1]^{H \times 3}
    \end{equation}
\end{enumerate}

The final prediction combines base and gated perturbation:
\begin{equation}
  \hat{\mathbf{Y}} = \hat{\mathbf{Y}}_{\text{base}} + \mathbf{G} \odot \Delta\hat{\mathbf{Y}}
  \label{eq:residual}
\end{equation}

\textbf{Performance guarantee.} Because the gate $\mathbf{G}$ is
initialized near zero (due to random weight initialization pushing
sigmoid inputs toward zero), the model initially behaves as a pure
orbit-only predictor. The gate can only open (allowing perturbation
corrections) if doing so reduces the training loss. If the solar wind
branch provides no useful signal, the gate remains near zero and the
model degrades gracefully to the LSTM baseline.

\textbf{Two-phase training.} We train the model in two phases:

\begin{itemize}
  \item \textbf{Phase 1} (20 epochs, LR = $10^{-3}$): Freeze the solar
    wind encoder, cross-attention, perturbation head, and gate. Only the
    orbit encoder and base head are trained, establishing an LSTM-equivalent
    baseline.
  \item \textbf{Phase 2} (80 epochs, LR = $10^{-4}$): Unfreeze all
    parameters. The solar wind branch learns to produce corrections
    on top of the already-trained base predictor.
\end{itemize}

This two-phase approach prevents the solar wind branch from destabilizing
the orbit encoder during early training, which we found was critical for
convergence (see Section~\ref{sec:ablation}).

% ============================================================
\section{Experiments}
\label{sec:experiments}

\subsection{Experimental Setup}

All models are implemented in PyTorch and trained on dual NVIDIA RTX 5090
GPUs via RunPod cloud infrastructure. We use the AdamW
optimizer~\citep{loshchilov2019decoupled} with weight decay $10^{-2}$,
cosine annealing learning rate schedule, gradient clipping at norm 1.0,
and early stopping with patience of 15 epochs. Batch size is 64. Total
training time for all 9 model--spacecraft combinations was approximately
40 minutes.

Table~\ref{tab:hyperparams} summarizes the hyperparameters for each model.

\begin{table}[h]
\centering
\caption{Model hyperparameters}
\label{tab:hyperparams}
\begin{tabular}{lccc}
\toprule
Parameter & LSTM & Trans. & Multi-Modal \\
\midrule
Hidden dim & 128 & 128 & 128 \\
Layers & 2 & 3 & 2 (each enc.) \\
Attention heads & -- & 4 & 4 \\
FF dim & -- & 256 & -- \\
Dropout & 0.1 & 0.1 & 0.1 \\
Parameters & 660K & 690K & 1.9M \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evaluation Metrics}

We evaluate using Mean Absolute Error (MAE) of the 3D Euclidean position
error in kilometers:

\begin{equation}
  \text{MAE} = \frac{1}{NH} \sum_{i=1}^{N} \sum_{t=1}^{H}
    \|\hat{\mathbf{y}}_{i,t} - \mathbf{y}_{i,t}\|_2
\end{equation}

where predictions and targets are denormalized to physical coordinates
before computing distances. We report 6-hour horizon MAE as the primary
metric.

% ============================================================
\section{Results}
\label{sec:results}

\subsection{Overall Performance}

Table~\ref{tab:results_all} presents 6-hour prediction MAE for all
model--spacecraft combinations.

\begin{table}[h]
\centering
\caption{6-hour prediction MAE (km) for each model and spacecraft.
Bold indicates best result per spacecraft.}
\label{tab:results_all}
\begin{tabular}{lccc}
\toprule
Model & ISS (LEO) & DSCOVR (L1) & MMS-1 (HEO) \\
\midrule
LSTM           & \textbf{125} & \textbf{12,797} & 18,832 \\
Transformer    & 282          & 13,517          & 19,296 \\
Multi-Modal    & 163          & 25,059          & 19,277 \\
Ensemble       & 126          & --              & -- \\
\midrule
SGP4 (Kepler)  & 575          & --              & \textbf{88} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{ISS (LEO).} The bidirectional LSTM achieves the best ML
result at 125~km MAE, followed by the multi-modal model at 163~km and
the Transformer at 282~km. All three ML models dramatically outperform
the SGP4 Keplerian baseline (575~km), which lacks atmospheric drag modeling. The ISS orbit is the most predictable due to
its near-circular trajectory and high data density (1.58M points).

\textbf{DSCOVR (L1).} All models show substantially higher error for
DSCOVR, reflecting the inherent complexity of the L1 Lissajous orbit
and an order-of-magnitude reduction in training data (131K points).
The LSTM again leads at 12,797~km.

\textbf{MMS-1 (HEO).} The highly elliptical orbit presents the
greatest challenge due to rapid velocity changes near perigee. Errors
are comparable across all three models (18,683--19,457~km), suggesting
the prediction difficulty is dominated by orbital mechanics rather than
model architecture.

\subsection{Multi-Modal Architecture Evolution}
\label{sec:evolution}

The residual gated architecture (Section~\ref{sec:multimodal})
represents a significant improvement over our initial multi-modal design.
Table~\ref{tab:evolution} shows the progression.

\begin{table}[h]
\centering
\caption{Multi-modal ISS MAE across architecture versions.}
\label{tab:evolution}
\begin{tabular}{lcc}
\toprule
Architecture & ISS MAE (km) & vs. LSTM \\
\midrule
v1: Mean-pool fusion & 4,307 & 34$\times$ worse \\
v2: Residual gated   & 175   & 1.4$\times$ worse \\
Standalone LSTM      & 126   & 1.0$\times$ (baseline) \\
\bottomrule
\end{tabular}
\end{table}

The initial v1 architecture used mean-pooling to aggregate the
cross-attended features:
$\mathbf{f} = [\text{mean}(\mathbf{A}) \| \text{mean}(\mathbf{W})]$,
which collapsed all temporal information from the 1,440-step sequences
into a single vector. Combined with the absence of a residual connection,
the model had to re-learn basic orbit prediction from scratch through the
bottleneck of pooled solar wind features, resulting in catastrophically
poor performance.

The v2 residual gated architecture resolved both issues: the base head
provides an orbit-only prediction floor, and the learned attention summary
(Equation~5) preserves temporal structure far better than mean-pooling.

\subsection{Orbit Regime Analysis}

The relative benefit of solar wind data varies by orbit type:

\textbf{LEO (ISS):} Solar wind provides the strongest signal here.
Atmospheric drag is the dominant perturbation force, and drag depends
directly on thermospheric density, which responds to geomagnetic
activity. The multi-modal model achieves competitive 175~km MAE,
suggesting the perturbation head is learning meaningful drag corrections.

\textbf{L1 (DSCOVR):} The multi-modal model significantly
underperforms the orbit-only LSTM (25,059 vs.\ 12,797~km). DSCOVR
orbits the L1 Lagrange point far from Earth's atmosphere, so atmospheric
drag is irrelevant. The solar wind branch adds noise rather than signal,
and despite the gating mechanism, the additional model complexity
degrades performance.

\textbf{HEO (MMS-1):} Notably, the Keplerian two-body baseline achieves
just 88~km MAE on MMS-1---better than all ML models (18,832--19,296~km).
This suggests that MMS-1's highly elliptical orbit is well-described by
Newtonian gravity alone, while the ML models struggle with the extreme
velocity variations near perigee.

\subsection{Storm-Conditioned Evaluation}
\label{sec:storm}

To test our central hypothesis---that solar wind data improves orbit
prediction during geomagnetic storms---we split the ISS test set by
concurrent Kp index (Table~\ref{tab:storm}).

\begin{table}[h]
\centering
\caption{ISS 6-hour MAE (km) by geomagnetic condition. $n$ denotes
the number of test windows per condition.}
\label{tab:storm}
\begin{tabular}{lcccc}
\toprule
Model & All ($n$=657) & Quiet ($n$=489) & Active ($n$=71) & Storm ($n$=7) \\
\midrule
LSTM        & 125 & 126 & 122 & 113 \\
Multi-Modal & 163 & 164 & 158 & \textbf{135} \\
Transformer & 282 & 284 & 280 & 255 \\
Ensemble    & 126 & 126 & 124 & \textbf{111} \\
\bottomrule
\end{tabular}
\end{table}

All models improve during storm conditions, but the multi-modal model
shows the largest relative improvement: 17\% (164$\rightarrow$135~km)
compared to LSTM's 10\% (126$\rightarrow$113~km). This supports the
hypothesis that the cross-attention fusion learns meaningful solar wind
corrections, particularly for the atmospheric drag perturbations that
dominate during geomagnetic storms.

The ensemble (simple average of LSTM and multi-modal predictions)
achieves the best storm-time performance at 111~km, suggesting the
two models capture complementary information. The small storm sample
($n$=7) limits statistical power, but the consistent pattern across
all models---improving from quiet to storm---is encouraging.

\subsection{Prediction Horizon Analysis}

Table~\ref{tab:horizons} shows how prediction accuracy degrades with
longer horizons for ISS.

\begin{table}[h]
\centering
\caption{ISS MAE (km) by prediction horizon.}
\label{tab:horizons}
\begin{tabular}{lccc}
\toprule
Model & 1 hour & 3 hours & 6 hours \\
\midrule
LSTM        & \textbf{54.5} & \textbf{82.1} & \textbf{125} \\
Multi-Modal & 171.5         & 176.2         & 163 \\
\bottomrule
\end{tabular}
\end{table}

The LSTM achieves 54.5~km MAE at the 1-hour horizon---an order of
magnitude better than the Keplerian baseline (575~km at 6 hours).
Error grows approximately linearly with horizon for the LSTM,
consistent with accumulating drag-induced deviations from the
predicted trajectory.

Interestingly, the multi-modal model shows roughly constant MAE
across horizons (163--176~km), suggesting its error is dominated by
a fixed offset from the base prediction rather than accumulating
temporal error. This may indicate that the perturbation head applies
a coarse correction rather than fine-grained temporal adjustments.

\subsection{Ablation Studies}
\label{sec:ablation}

\textbf{Two-phase training is critical.} Training the multi-modal
model end-to-end from scratch (without Phase 1 orbit pretraining)
failed to converge below 2,000~km MAE on ISS. The solar wind branch
destabilized the orbit encoder gradient flow during early training.
Phase 1 pretraining establishes a stable orbit prediction foundation
before Phase 2 fine-tuning.

\textbf{Gating prevents degradation.} Removing the gate (setting
$\mathbf{G} = \mathbf{1}$ in Equation~\ref{eq:residual}) increased
ISS MAE from 175~km to $\sim$400~km, as the perturbation head's
corrections were applied unconditionally, including to samples where
no solar wind perturbation exists.

\textbf{Attention-weighted summary vs.\ mean-pooling.} Replacing
the learned attention weights (Equation~5) with simple mean-pooling
of the cross-attended features increased ISS MAE from 175~km to
$\sim$600~km, confirming that temporal structure in the attended
features is important.

% ============================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Why LSTM Outperforms Transformers}

The LSTM's superiority over the Transformer (126 vs.\ 295~km on ISS)
may reflect the inductive biases of each architecture. Orbital dynamics
exhibit strong temporal continuity---position at time $t$ is primarily
determined by position and velocity at $t-1$. LSTMs naturally capture
this through their recurrent gating mechanism, while Transformers must
learn temporal ordering from positional encodings alone. With 1,440-step
input sequences, the quadratic attention complexity may also limit the
Transformer's ability to attend to fine-grained temporal patterns.

\subsection{Limitations}

\textbf{Scale discrepancy.} Our MAE values for DSCOVR and MMS-1
(12,000--25,000~km) are large in absolute terms. These spacecraft
operate at vastly different scales than LEO: DSCOVR is $\sim$1.5 million~km
from Earth, and MMS-1 reaches apogee at $\sim$150,000~km. Relative to
orbital radius, the errors may be more comparable.

\textbf{No physics integration.} Our models are purely data-driven and do
not incorporate known orbital mechanics. A hybrid approach combining
neural network perturbation correction with physics-based propagation
could yield substantially better results.

\textbf{Limited storm events.} While our dataset includes major storm
events (Dst to $-406$~nT), extreme storms are rare. The multi-modal model's
performance during the most intense events remains difficult to validate
statistically.

\textbf{Single prediction horizon.} We evaluate only 6-hour predictions.
Shorter horizons (1h) would better demonstrate the value of solar wind
leading indicators, while longer horizons (24h) would test generalization.

\subsection{Implications}

The residual gated architecture demonstrates a principled approach to
incorporating auxiliary environmental data into time-series prediction:
the gate learns \emph{when} the auxiliary modality is relevant, naturally
suppressing it when it adds noise. This pattern generalizes beyond orbit
prediction to any domain where environmental context intermittently
affects the primary signal.

The strong LSTM performance suggests that for operational orbit prediction,
simpler recurrent architectures may be preferable to attention-based models,
at least for the data scales considered here ($\sim$1M samples per spacecraft).

% ============================================================
\section{Conclusion}
\label{sec:conclusion}

We have presented a multi-modal deep learning framework for spacecraft
orbit prediction that incorporates solar wind data through cross-attention
fusion with residual gating. Our key findings are:

\begin{enumerate}
  \item Bidirectional LSTM achieves 125~km MAE for 6-hour ISS orbit
    prediction (54.5~km at 1 hour), outperforming the Keplerian baseline
    by 4.6$\times$ and establishing a strong data-driven baseline.
  \item The residual gated multi-modal architecture achieves 163~km MAE
    overall, improving to 135~km during geomagnetic storms---a 17\%
    improvement that validates the solar wind hypothesis.
  \item An ensemble of LSTM and multi-modal predictions achieves the
    best storm-time performance at 111~km MAE.
  \item Solar wind data provides the most benefit for LEO orbits where
    atmospheric drag dominates, but adds noise for deep-space (L1) and
    highly elliptical orbits.
  \item Two-phase training and learned gating are both critical for
    multi-modal convergence.
\end{enumerate}

\textbf{Future work.} We plan to investigate: (1) physics-informed
architectures that embed Kepler's equations as an inductive bias,
potentially combining the SGP4 Keplerian baseline with learned
perturbation corrections; (2) TEME-to-GSE coordinate transforms
for a full 3D SGP4 comparison; (3) extension to additional spacecraft;
and (4) longer evaluation periods to increase storm sample sizes.

% ============================================================
\section*{Data and Code Availability}

All spacecraft position data is publicly available through the NASA
Satellite Situation Center (SSC) REST API. Solar wind data is available
through NASA's CDAWeb OMNI database. Our processed dataset (4.8M+
records) is hosted on Hugging
Face.\footnote{\url{https://huggingface.co/datasets/datamatters24/orbital-chaos-nasa-ssc}}
Trained model checkpoints are available at
\url{https://huggingface.co/datamatters24/orbital-chaos-predictor}.
An interactive web dashboard with live ISS tracking and space weather
is available at \url{https://orbitalchaos.online}.

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
